Formal verification of train route planning

### 2025-06-27 paper/main.tex
corrections !
corrections

\begin{document}
%
\title{Formal verification of train route planning}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Lucas Villaume \and Guillaume Bonfante \and Martin Vassor}
%
\authorrunning{L. Villaume et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Université de Lorraine, CNRS, Inria, LORIA, F-54000 Nancy, France}
%
\maketitle
%
\begin{abstract}
	% Contexte
	Trains networks are critical system which failure may lead to numerous fatalities, in case of crashes, or important loss, if trains are not properly routed. 
	% Problème
	In this paper, our goal is to formally verify the correctness of train routes. More precisely, we verify the absence of deadlocks, the termination, and the absence of crashes.
	% Approche
	To address this question, we model train systems on paper and in TLA+. Our model includes trains, as well as switches, detection sensors and traffic lights. Both switches and traffic lights are controlled by a regulator upon train events. We provide trains and the regulator with orders, characterising the schedule, which is then checked using TLA+'s model checker.
	% Resultats
	Our contributions are threefold: first, we provide formal model of railroad networks and schedules; second, we implement this model as a TLA+ specification and we report tests on checking schedules on a non-trivial network, composed of 3 trains, 8 blocks and 5 switches; finally we show how to sequentially compose schedules, which we check against our TLA+ specification.

@@

\paragraph{Routing trains.}
\todo{Ajouter: tout est automatique}
Routes are computed on the fly. The schedule is renewed every "day"\footnote{We use a time scale factor: one of our "day" corresponds to 20 minutes in real time.}  and we introduce some randomness in the process: beyond some routine daily planning, some trains are inserted/deleted, some trains departure/destination may change, and so on. We simulate also delays. Finally, in our model, we distinguish freight and  passengers routes giving priority to the latter. Routes are recomputed each time a train has finished its duty. The general motivation of this paper is to propose a lightweight scheduling solution that is adapted to the fully automatic control, that takes into account the distributed nature of the system, that is reliable and does not suffer from interlocking.

@@

Fourth,  the regulator is capable of reading train block entrance events, requiring some switch swap and finally locking (that is make sure a signal is red)/unlocking some signal. We consider that those latter operations can be carried out without delay. Finally, the regulator may append some new duty to a train. To sum up, to control the train mouvements, the regulator can only lock/unlock signals. Naturally, a signal intended to block train $T_1$ may block $T_2$ if not set at the right time.

 Parler de la stratégie, du modèle générale. All these operations are done by what we call  \emph{orders}. Roughly speaking, trains Orders have the shape "Start in direction Until position". Regulator orders are: \todo{en parler}.

 To sum up, the regulation policy can be seen as an asynchronous distributed system. The N trains and the regulator are independent agents. That led us to TLA+, a framework introduced by Leslie Lamport for these kind of problems~\cite{Lamport}.

\todo{ Comparaison avec les travaux antérieurs / lit review}

Due to the distributed nature of the regulation, building a safe policy can be very tricky. We want to show in this paper that tools from formal methods are very useful in this respect.  In this paper, we describe in  a first step the formalisation process. We go up to a TLA+ specification that will allow us checking two properties of route orchestration: safety and liveness. This is our first research question.

@@

For our last research problem, we show that TLA+ may be used to refine a regulation policy. In our first attempts, we had typically some deadlocks issues. We have been able to detect them via TLA+ and then to provide some correction.

@@

\paragraph{Model.}
Before delving further in the details, we introduce a real-world N-scale train model (\ref{fig:maquette}), which shaped our approach and was used to test our results. This model has 8 blocks, 5 switches. Each block is bidirectional with a signal at each end. Blocks size goes from 25cm to 70cm. We operate three trains concurrently. Due to the size of the model, we reduced our signals  to "red" and "green" status, the trains being too close to each other for a finer modelling of signals. Furthermore, in case of a red signal, we make the hypothesis that the train are able to stop before reaching the end of the block.

@@

Given some set $S$ and some $x \not\in S$, we denote by $S_x = S \cup \{ x\}$.  An other notation: given some function $f: X \to Y$, $x \in X$ and $y \in Y$, we denote by $f[x \leftarrow y]$, the function:
 $$f[x \leftarrow y] \triangleq \left\{  \begin{array}{ll} x &\mapsto y\\
x' \neq x \in X &\mapsto f(x')
\end{array}\right.$$

@@

Our model is has $k+1$ agents corresponding to the $k$ trains and the Regulator. Those agents take place on an infrastructure consisting of a track network equipped with signals. We first present the infrastructure model, then the agents.

@@

We suppose that the train network can be described as follows. First, we consider a set of $n$ blocks: $\blocks = \{ \bid{1}, \bid{2}, \ldots, \bid{n}\}$. Blocks are oriented: there is a $\dirForward$ (forward) end and a $\dirBackward$ (backward) one. Let $\directions = \{\dirForward, \dirBackward\}$ be the set of directions.  Second, we consider a set of $m$ trackworks\footnote{For the sake of simplicity, in this paper, we only consider switches. Our work can easily be extended to crossings and double slip switches (e.g. crossings have only one internal state and double slip switches have four states).}: $\turnouts = \{\sid{1}, \ldots, \sid{m}\}$. Trackworks are characterized by an internal state: switches can be in direct or deviate position (next denoted '\direct' and '\deviate'). Let $\internalState$ denotes the set of internal states. A \emph{network configuration} is given by a function $\sigma: \turnouts \to \internalState$. Let $\networkConf$ denotes the set of all network configurations. For instance, for two turnouts $\theta = \{\sid{1}, \sid{2}\}$, the function $\sid{1} \mapsto \deviate, \sid{2} \mapsto \direct$ states that switch $\sid{1}$ is in deviate position while \sid{2} is in direct one. The latter function is also described via an array  $[\deviate, \direct]$ giving states in trackwork order.

The network is then described as a triple $(\blocks, \turnouts, \sucblock)$ where $\sucblock$ is a partial function indicating the following block given a starting position, a direction and a network configuration: $\sucblock: \blocks\times \directions \times \networkConf \to \blocks$. We note $\suc{\bid{id}}{d}{\switches} = \nosuc$ when it is undefined. From an abstract point of view, the topology of the network reduces to that data.

\paragraph{Signals.}
\todo{Est-ce qu'on met les signaux dans les agents ?}
At each end of the blocks, there is a signal whose state is \sigred (red) or \siggreen (green). In other words, the status of signals is described by a function $\signals: \blocks \times \directions \to \{ \sigred, \siggreen\}$. The signal status is determined by the network configuration and explicit locking requests. Given some signal status $\signals$, a block $\bid{id}$ and a direction $\dirFmt{d}$, $\signalF{\bid{id}}{d} = \sigred$ means the signal guarding the exit of block $\bid{id}$ in direction $\dirFmt{d}$ is red, and we expect trains not leaving that block going in this direction\footnote{Our semantics, presented in Section~\ref{sec:model_semantics}, ensure such behaviour.}. We note $\setSignalsTo{\bid{\text{tgt}}}{c}$ the signal state $\signals$ except signals of block \bid{\text{tgt}}, which are turned to color $c$: 
$$(\setSignalsTo{\bid{\text{tgt}}}{c})(\bid{\text{from}}, \dirFmt{d}) = \begin{cases}
	c & \text{if }\bid{\text{from}} = \bid{\text{tgt}}\\
	\signalF{\bid{\text{from}}}{d} & \text{otherwise}
\end{cases}$$

@@

Thus, each train is a quadruple $\trainFmt{t} = \trainTuple{\tid{id}}{\bid{id}}{d}{P}$ with $\tid{id}$ the train's identifier, $\bid{id}$ its position, $\dirFmt{d}$ its direction and $P$ its sequence of orders. Trains are collected in a sequence $\trainSeq = (\trainFmt{t_1}, \ldots, \trainFmt{t_k})$ where each $\trainFmt{t_i}$ being the tuple of train $\trainFmt{i}$.

\paragraph{The Regulator.}
The regulator is in charge of the execution of the routing.
It is characterised by a routing program, a token list, and a waiting set.

\begin{itemize}
	\item The routing program $E$ specifies how the regulator should react to events (i.e. when a train enters a block). A program is a collection of rules to apply upon each event (called an \emph{event handler}). We detail further event handlers below.
	\item The token list, denoted $\tokens$, is a function  $\tokens: \blocks \to \Nat$ associating a token level to each block. Those tokens are used to prevent multiple trains accessing the same block at the same time.
	\item The waiting set is a set $W \subseteq \blocks \times \blocks \times \Nat \times \trains$. A 4-tuple $\tuple{\bid{src}, \bid{dst}, n, \tid{id}} \in W$ means that train \tid{id} is currently waiting in the block \bid{src} for $\tokenOf{\bid{dst}}$ to be increased up to $n$.
\end{itemize}

Overall, the regulator is a tuple $R = \regTuple{E}{\tokens}{W}$. We note $\incrToken{T}{\bid{id}}$ the list of tokens identical to $T$, except for $\tokenOf{\bid{id}}$, which is incremented.

@@

Train events are handled by the regulator. Upon such events, the regulator can take actions such as turning a switch, allowing or blocking a train, etc.. The regulator is equipped with a routing program. This routing program contains, for each train, a list of handlers. A handler is a tuple $\tuple{\bid{id}, A}$, where $\bid{id}$ is the position the train occupies when triggering the event, and $A$ is the list of orders associated with that event. For a given train, the first handler of the list is the set of actions that have to be executed on the train's first event, second handler for the second event, and so on.
Given a routing program $E$, we note $\handlerOf{E}{\tid{id}}$ the list of handlers of train $\tid{id}$, $\head{\handlerOf{E}{\tid{id}}} = \handlerOf{E}{\tid{id}}[1]$ the first handler of the list. Given a handler $h = \tuple{\bid{id}, A}$, we note $\actionsOf{h}$ (resp. \blockOf{h}) the list of action (resp. the block id) of $h$. We note $\nextAct{\handlerOf{E}{\tid{id}}}$ the routing program $E$ where the first event handler of train $\tid{id}$ is replaced with $\handler{\blockOf{\head{\handlerOf{E}{\tid{id}}}}}{\buftail{\actionsOf{\head{\handlerOf{E}{\tid{id}}}}}}$, i.e. where the first action of the first handler of train \tid{id} is removed. We note $\popHandlerHead{E}{\tid{id}}$ the routing program $E$ where the first event handler of train \tid{id} is removed, i.e. where $\handlerOf{E}{\tid{id}}$ is replaced by $\buftail{\handlerOf{E}{\tid{id}}}$.

Actions contained in a handler are of the four following kinds:
\begin{description}
	\item [\wait{\bid{id}}{n}:] the train waits until the token of block \bid{id} reaches value $n$
	\item [\incr{\bid{id}}:] this increments the token $\bid{id}$ in $T$, thus releasing a train waiting
	\item [\turn{\sid{id}}{s}:] this turns the switch $\sid{id}$ into the $s$ state: $\sigma\cdot\turn{\sid{id}}{s} = \sigma[\sid{id} \leftarrow s]$
	\item [\auth:] a shortcut for waiting on oneself
\end{description}

\paragraph{Model.}
Overall, states of our model are tuples $\tuple{\trainSeq, \regulator, \switches, \signals}$ where $\trainSeq$ is a set of train tuples, $\regulator$ a regulator tuple, $\switches$ a network configuration (indicating trackworks states) and $\signals$ a signals state function. We suppose all train tuples in $\trainSeq$ have a different identifier. Notice that $\sucblock$ is not part of the state. Instead, the dynamics of the model (see below) is parameterised by $\sucblock$. Let $\modelSet$ be the set of all model states.

@@

In the previous section, we introduced our model. In this section, we present the dynamics of our model, that is the rules that govern how our model can switch from one state to the other.
We enrich our model with two elements: a guard, which goal is to schedule agents (for instance, it ensures the regulator acts atomically); and a communication buffer, which stores pending events until the regulator treats them. We present those two new elements in the first subsection.
The dynamics is described by means of a reduction relation, characterising the successive states of our model. We define this reduction relation using inference rules in the second subsection.

\todo{Expliquer le temps discret et le mix grand pas/petit pas.}

@@

In our dynamic rules, we distinguish three kinds of rules: one kind for trains, one kind for the regulator, and one kind for updating signals. We intend the signals' and regulator's reactions to be executed atomically. On the other hand, trains evolve much slower. To reflect this behaviour in our rules, we introduce a \emph{guard}. While, in our rules, the reaction of the regulator can take several steps, to implement atomic reaction, the guard blocks the rest of the system. Similarly, upon any train movement or regulator reaction, signals can be updated. In practice, such update happens instantly. To mimick this reaction of signal w.r.t. other events, the guard is in charge of blocking the system until signals are up-to-date.

Essentially, this guard is the state machine shown in Figure~\ref{fig:state_machine_guard}. The four states of the guard are \guardR, \guardS, and \guardT.

\begin{figure}
	\centering
	\begin{tikzpicture}
		\node[draw] (trains) {Train movements (\guardT)};
		\node[left=.5cm of trains.west] (zero) {};
		\node[draw, right=of trains] (regulator) {Regulator reaction (\guardR)};
		\node[draw, below=of regulator] (signals) {Signals update (\guardS)};

@@

In our model, a train event represents a train going from one block to another. Train events are not dealt with instantly by the regulator (e.g. two train events can happen before the regulator begins to react to the first one). To characterise this behaviour, train events are stored in a buffer until the regulator uses them. When such thing happens, the train sends a message with its identifier to the regulator. A \emph{train buffer} \bufTrain is a sequence of such events, i.e. of train identifiers.
Similarly, the regulator communicates with signals with a \emph{signal buffer} \bufSig, which is a list of tuples $\tuple{\bid{id}, c}$, indicating that both signals of block $\bid{id}$ should be turned to value $c$. Let \bufTrainSet (resp. \bufSigSet) denote the set of train (resp. signal) buffers.

@@

States of the system are 4-tuples from $\{\guardR, \guardS, \guardT\} \times \bufTrainSet \times \bufSigSet \times \modelSet$. Our reduction relation is a binary relation on system states, noted $\reduces$: $S_1 \reduces S_2$ means that state $S_1$ can change to $S_2$ (note that this relation is not a function: $S_1$ can reduce to multiple states, due to non-determinism). We define $\reduces$ as the least relation induced by the inference rules presented hereafter.

@@

\begin{mathpar}
	\inferrule*[left=\ruleDef{Start}]{
		\dirFmt{d} \neq \dirFmt{d^\prime} \and P = \trainConcat{\su{d^\prime}{\posFmt{N}}}{P^\prime}
	}{
		\hspace{3mm} \redTuple{\guardT}{\bufTrain}{\bufSig}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id}}{d}{P}\}}{\regulator}{\switches}{\signals}} \\
		\reduces
		\redTuple{\guardT}{\push{\bufTrain}{\tid{id}}}{\bufSig}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id}}{d^\prime}{P}\}}{\regulator}{\switches}{\signals}}
	}
\end{mathpar}

Conversely, if the train program is empty (i.e. upon termination), the train stops using rule \ruleRef{Stop}.
\begin{mathpar}
	\inferrule*[left=\ruleDef{Stop}]{
		\dirFmt{d} \neq \dirStop
	}{
	\hspace{4mm}  \redTuple{\guardT}{\bufTrain}{\bufSig}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id}}{d}{\emptyTrainProg}\}}{\regulator}{\switches}{\signals}} \\
	\reduces
	\redTuple{\guardT}{\push{\bufTrain}{\tid{id}}}{\bufSig}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id}}{\dirStop}{\emptyTrainProg}\}}{\regulator}{\switches}{\signals}}
	}
\end{mathpar}

The last two rules for trains describe what happens when a train actually moves, i.e. when its position changes. Two cases can occur, depending on the list of intermediate blocks $\posFmt{N}$ in the current order of the train
If $\posFmt{N}$ contains only the upcoming block, then upon leaving the block, the order is removed (rule \ruleRef{UntilNext}); otherwise the upcoming block is simply popped from $\posFmt{N}$ (rule \ruleRef{Until}). Notice that in both cases, the state of the guard changes to \guardS, meaning that signals will be updated immediately after that rule. Finally, when the train reaches intermediate positions (rule \ruleRef{Until}), a train event is enqueued for the regulator.

\begin{mathpar}
	\inferrule*[left=\ruleDef{Until}] {
		\suc{\bid{id}}{d}{\switches} = \bid{id^\prime} \neq \nosuc
		\and
		\signalF{\bid{id}}{d} = \siggreen
		\and
		\posFmt{N}\neq \varepsilon
	}{
		\redTuple{\guardT}{\bufTrain}{\bufSig}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id}}{d}{\trainConcat{\su{d}{[\bid{id^\prime}; \posFmt{N}]}}{P}}\}}{\regulator}{\switches}{\signals}} \\
		\reduces		
		\redTuple{\guardS}{\push{\bufTrain}{\tid{id}}}{\bufSig}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id^\prime}}{d}{\trainConcat{\su{d}{[\posFmt{N}]}}{P}}\}}{\regulator}{\switches}{\signals}}
	}
\end{mathpar}

>>>>

\begin{document}
%
\title{Formal verification of train route planning}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Lucas Villaume \and Guillaume Bonfante \and Martin Vassor}
%
\authorrunning{L. Villaume et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Université de Lorraine, CNRS, Inria, LORIA, F-54000 Nancy, France}
%
\maketitle
%
\begin{abstract}
	% Contexte
	Trains networks are critical system which failure may lead to numerous fatalities, in case of crashes, or important loss, if trains are not properly routed. 
	% Problème
	In this paper, our goal is to formally verify the correctness of train routes. More precisely, we verify the absence of deadlocks, the termination, and the absence of crashes.
	% Approche
	To address this question, we model train systems on paper and in TLA+. Our model includes trains, as well as switches, detection sensors and traffic lights. 
	%Both switches and traffic lights are controlled by a regulator upon train events. 
	We provide trains and the regulator with orders, characterising the schedule, which is then checked using TLA+'s model checker.
	% Resultats
	Our contributions are threefold: first, we provide formal model of railroad networks and schedules; second, we implement this model as a TLA+ specification and we report tests on checking schedules on a non-trivial network, composed of 3 trains, 8 blocks and 5 switches; finally we show how to sequentially compose schedules, which we check against our TLA+ specification.

@@

\paragraph{Routing trains.}
%\todo{Ajouter: tout est automatique}
Routes are computed on the fly. The schedule is renewed every "day"\footnote{We use a time scale factor: one of our "day" corresponds to 20 minutes in real time.}  and we introduce some randomness in the process: beyond some routine daily planning, some trains are inserted/deleted, some trains departure/destination may change, and so on. We simulate also delays. Finally, in our model, we distinguish freight and  passengers routes giving priority to the latter. Routes are recomputed each time a train has finished its duty. The general motivation of this paper is to propose a lightweight scheduling solution that is adapted to the fully automatic control, that takes into account the distributed nature of the system, that is reliable and does not suffer from interlocking.

@@

Fourth,  the regulator is capable of reading train block entrance events, requiring some switch swap and finally locking (that is make sure a signal is red)/unlocking some signal. We consider that those latter operations can be carried out without delay. Finally, the regulator may append some new duty to a train. To sum up, to control the train mouvements, the regulator can only lock/unlock signals. Naturally, a signal intended to block train $t_1$ may block $t_2$ if not set at the right time.

% Parler de la stratégie, du modèle générale. All these operations are done by what we call  \emph{orders}. Roughly speaking, trains Orders have the shape "Start in direction Until position". Regulator orders are: %\todo{en parler}.

 To sum up, the regulation policy can be seen as an asynchronous distributed system. The N trains and the regulator are independent agents. That led us to TLA+, a framework introduced by Leslie Lamport for these kind of problems~\cite{Lamport}.  
%\todo{ Comparaison avec les travaux antérieurs / lit review}
Due to the distributed nature of the regulation, building a safe policy can be very tricky. We want to show in this paper that tools from formal methods are very useful in this respect. We describe In a first step, we describe a TLA+ specification that will allow us checking two properties of route orchestration: safety and liveness. This is our first research question.

@@

For our last research problem, we show how one can build a global planning by means of composition of elementary steps. We check with TLA+ those compositions are correct.

@@

\paragraph{Model.}
Before delving further in the details, we introduce a real-world N-scale train model (\ref{fig:maquette}), which shaped our approach and was used to test our results. This model has 8 blocks, 5 switches. Each block is bidirectional with a signal at each end. %Blocks size goes from 25cm to 70cm.
 We operate three trains concurrently. Due to the size of the model, we reduced our signals  to "red" and "green" status, the trains being too close to each other for a finer modelling of signals. Furthermore, in case of a red signal, we make the hypothesis that the train are able to stop before reaching the end of the block.

@@

Given some function $f: X \to Y$, $x \in X$ and $y \in Y$, we denote by $f[x \leftarrow y]$ the function from $X$ to $Y$ verifying $f[x \leftarrow y](x) = y$ and for any $z \neq x$, $f[x \leftarrow y](z) = f(z)$.

@@

Our model has $k+1$ agents corresponding to the $k$ trains and the Regulator. Those agents take place on an infrastructure consisting of a track network equipped with signals. We first present the infrastructure model, then the agents.

@@

We suppose that the train network can be described as follows. First, we consider a set of $n$ blocks: $\blocks = \{ \bid{1}, \bid{2}, \ldots, \bid{n}\}$. Blocks are oriented: there is a $\dirForward$ (forward) end and a $\dirBackward$ (backward) one. Let $\directions = \{\dirForward, \dirBackward\}$ be the set of directions.  Second, we consider a set of $m$ trackworks\footnote{For the sake of simplicity, in this paper, we only consider switches. Our work can  be extended to crossings and double slip switches (e.g. crossings have only one internal state and double slip switches have four states).}: $\turnouts = \{\sid{1}, \ldots, \sid{m}\}$. Trackworks are characterized by an internal state: switches can be in direct or deviate position (next denoted '\direct' and '\deviate'). Let $\internalState$ denotes the set of internal states. A \emph{network configuration} is given by a function $\sigma: \turnouts \to \internalState$. Let $\networkConf$ denotes the set of all network configurations. For instance, for two turnouts $\theta = \{\sid{1}, \sid{2}\}$, the function $\sid{1} \mapsto \deviate, \sid{2} \mapsto \direct$ states that switch $\sid{1}$ is in deviate position while \sid{2} is in direct one.
% The latter function is also described via an array  $[\deviate, \direct]$ giving states in trackwork order.

The network is then described as a triple $(\blocks, \turnouts, \sucblock)$ where $\sucblock$ is a partial function indicating the following block given a starting position, a direction and a network configuration: $\sucblock: \blocks\times \directions \times \networkConf \to \blocks$. We note $\suc{\bid{i}}{d}{\switches} = \nosuc$ when it is undefined. From an abstract point of view, the topology of the network reduces to that data.

\paragraph{Signals.}
%\todo{Est-ce qu'on met les signaux dans les agents ?}
At each end of the blocks, there is a signal whose state is \sigred (red) or \siggreen (green). In other words, the status of signals is described by a function $\signals: \blocks \times \directions \to \{ \sigred, \siggreen\}$. The signal status is determined by the network configuration and explicit locking requests. Given some signal status $\signals$, a block $\bid{i}$ and a direction $\dirFmt{d}$, $\signalF{\bid{i}}{d} = \sigred$ means the signal guarding the exit of block $\bid{i}$ in direction $\dirFmt{d}$ is red, and we expect trains not leaving that block going in this direction\footnote{Our semantics, presented in Section~\ref{sec:model_semantics}, ensure such behaviour.}. We note $\setSignalsTo{\bid{\text{tgt}}}{c}$ the signal state $\signals$ except signals of block \bid{\text{tgt}}, which are turned to color $c$: 
$$(\setSignalsTo{\bid{\text{tgt}}}{c})(\bid{\text{from}}, \dirFmt{d}) = \begin{cases}
	c & \text{if }\bid{\text{from}} = \bid{\text{tgt}}\\
	\signalF{\bid{\text{from}}}{d} & \text{otherwise}
\end{cases}$$

@@

Thus, each train is a quadruple $\trainFmt{{\bf t}} = \trainTuple{\tid{i}}{\bid{i}}{d}{P}$ with $\tid{i}$ the train's identifier, $\bid{i}$ its position, $\dirFmt{d}$ its direction and $P$ its sequence of orders. Trains are collected in a sequence $\trainSeq = (\trainFmt{{\bf t_1}}, \ldots, \trainFmt{{\bf t_k}})$ where each $\trainFmt{{\bf t_i}}$ being the tuple of train $\trainFmt{\tid{i}}$.

\paragraph{The Regulator.}
The regulator is in charge of the execution of the routing. It is characterised by a routing program, a token list, and a waiting set. Overall, the regulator is a tuple $R = \regTuple{E}{\tokens}{W}$.

\begin{itemize}
	\item The routing program $E$ specifies how the regulator should react to events (i.e. when a train enters a block). A program is a collection of rules to apply upon each event (called an \emph{event handler}). We detail further event handlers below.
	\item The token list, denoted $\tokens$, is a function  $\tokens: \blocks \to \Nat$ associating a token level to each block. Those tokens are used to prevent multiple trains accessing the same block at the same time.
	\item The waiting set is a set $W \subseteq \blocks \times \blocks \times \Nat \times \trains$. A 4-tuple $\tuple{\bid{src}, \bid{dst}, n, \tid{i}} \in W$ means that train \tid{i} is currently waiting in the block \bid{src} for $\tokenOf{\bid{dst}}$ to be increased up to $n$.
\end{itemize}

%We note $\incrToken{T}{\bid{i}}$ the list of tokens identical to $T$, except for $\tokenOf{\bid{i}}$, which is incremented.

@@

Train events are handled by the regulator. Upon such events, the regulator can take actions such as turning a switch, allowing or blocking a train, etc. The regulator is equipped with a routing program. This routing program contains, for each train, a list of handlers. A handler is a tuple $\tuple{\bid{i}, A}$, where $\bid{i}$ is the position the train occupies when triggering the event, and $A$ is the list of orders associated with that event. For a given train, the first handler of the list is the set of actions that have to be executed on the train's first event, second handler for the second event, and so on.
Given a routing program $E$, we note $\handlerOf{E}{\tid{i}}$ the list of handlers of train $\tid{i}$, $\head{\handlerOf{E}{\tid{i}}} = \handlerOf{E}{\tid{i}}[1]$ the first handler of the list. Given a handler $h = \tuple{\bid{i}, a \cdot A}$, let  $\actionsOf{h} \triangleq a$, $\blockOf{h} \triangleq \bid{i}$ and $\nextAct{h} \triangleq \tuple{\bid{i}, A}$.  %(resp. \blockOf{h}) the list of action (resp. the block id) of $h$.
% We note $\nextAct{\handlerOf{E}{\tid{i}}}$ the routing program $E$ where the first event handler of train $\tid{i}$ is replaced with $\handler{\blockOf{\head{\handlerOf{E}{\tid{i}}}}}{\buftail{\actionsOf{\head{\handlerOf{E}{\tid{i}}}}}}$, i.e. where the first action of the first handler of train \tid{i} is removed.
 We note $\popHandlerHead{E}{\tid{i}}$ the routing program $E$ where the first event handler of train \tid{i} is removed, i.e. where $\handlerOf{E}{\tid{i}}$ is replaced by $\buftail{\handlerOf{E}{\tid{i}}}$.

Actions contained in a handler are of the four following kinds:
\begin{description}
	\item [\wait{\bid{i}}{n}:] the train waits until the token of block \bid{i} reaches value $n$
	\item [\incr{\bid{i}}:] this increments the token $\bid{i}$ in $T$, thus releasing a train waiting
	\item [\turn{\sid{i}}{s}:] this turns the switch $\sid{i}$ into the $s$ state: $\sigma\cdot\turn{\sid{i}}{s} = \sigma[\sid{i} \leftarrow s]$
	\item [\auth:] a shortcut for waiting on oneself
\end{description}

\paragraph{Model.}
Overall, states of our model are tuples $\tuple{\trainSeq, \regulator, \switches, \signals}$ where $\trainSeq$ is a set of train tuples, $\regulator$ a regulator tuple, $\switches$ a network configuration (indicating trackworks states) and $\signals$ a signals state function.
% We suppose all train tuples in $\trainSeq$ have a different identifier. 
 Notice that $\sucblock$ is not part of the state. Instead, the dynamics of the model (see below) is parameterised by $\sucblock$. Let $\modelSet$ be the set of all model states.

@@

%In the previous section, we introduced our model. In this section, 
We present now the dynamics of our model, that is the rules that govern how our model can switch from one state to the other.
We enrich our model with three elements: a guard and two buffers. They ensure agents' autonomy. %, which goal is to schedule agents (for instance, it ensures the regulator acts atomically); and a communication buffer, which stores pending events until the regulator treats them. 
We present those two new elements in the first subsection.
The dynamics is described by means of a reduction relation, characterising the successive states of our model. We define this reduction relation using inference rules in the second subsection.

%\todo{Expliquer le temps discret et le mix grand pas/petit pas.}

@@

In our dynamic rules, we distinguish three kinds of rules: one kind for trains, one kind for the regulator, and one kind for  signals. We intend the signals' and regulator's reactions to be executed atomically. On the other hand, trains evolve much slower. To reflect this behaviour in our rules, we introduce a \emph{guard}. While, in our rules, the reaction of the regulator can take several steps, to implement atomic reaction, the guard blocks the rest of the system. Similarly, upon any train movement or regulator reaction, signals can be updated. In practice, such update happens instantly. To mimick this reaction of signal w.r.t. other events, the guard is in charge of blocking the system until signals are up-to-date.

Essentially, this guard is the state machine shown in Figure~\ref{fig:state_machine_guard}. The three guard states are $Q = \{\guardR, \guardS, \guardT\}$.
\begin{figure}
	\centering
	\begin{tikzpicture}
		\node[draw] (trains) {Train movements (\guardT)};
		\node[left=.5cm of trains.west] (zero) {};
		\node[draw, right=of trains] (regulator) {Regulator reaction (\guardR)};
		\node[draw, right=of regulator] (signals) {Signals update (\guardS)};

@@

In our model, a train event represents a train going from one block to another. Train events are not dealt with instantly by the regulator (e.g. two train events can happen before the regulator begins to react to the first one). To characterise this behaviour, train events are stored in a buffer until the regulator uses them. When such thing happens, the train sends a message with its identifier to the regulator. A \emph{train buffer} \bufTrain is a sequence of such events, i.e. of train identifiers.
Similarly, the regulator communicates with signals with a \emph{signal buffer} \bufSig, which is a list of tuples $\tuple{\bid{i}, c}$, indicating that both signals of block $\bid{i}$ should be turned to value $c$. Let \bufTrainSet (resp. \bufSigSet) denote the set of train (resp. signal) buffers.

@@

States of the system are 4-tuples from $Q\times \bufTrainSet \times \bufSigSet \times \modelSet$. Our reduction relation is a binary relation on system states, noted $\reduces$: $S_1 \reduces S_2$ means that state $S_1$ can change to $S_2$ (note that this relation is not a function: $S_1$ can reduce to multiple states, due to non-determinism). We define $\reduces$ as the least relation induced by the inference rules presented hereafter.

@@

\begin{mathpar}
	\inferrule*[left=\ruleDef{Start}]{
		\dirFmt{d} \neq \dirFmt{d^\prime} \and P = \trainConcat{\su{d^\prime}{\posFmt{N}}}{P^\prime}
	}{
		\hspace{3mm} \redTuple{\guardT}{\bufTrain}{\bufSig}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{m}}{\bid{i}}{d}{P}\}}{\regulator}{\switches}{\signals}} \\
		\reduces
		\redTuple{\guardT}{\push{\bufTrain}{\tid{m}}}{\bufSig}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{m}}{\bid{i}}{d^\prime}{P}\}}{\regulator}{\switches}{\signals}}
	}
\end{mathpar}

Conversely, if the train program is empty (i.e. upon termination), the train stops using rule \ruleRef{Stop}.
\begin{mathpar}
	\inferrule*[left=\ruleDef{Stop}]{
		\dirFmt{d} \neq \dirStop
	}{
	\hspace{4mm}  \redTuple{\guardT}{\bufTrain}{\bufSig}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{m}}{\bid{i}}{d}{\emptyTrainProg}\}}{\regulator}{\switches}{\signals}} \\
	\reduces
	\redTuple{\guardT}{\push{\bufTrain}{\tid{m}}}{\bufSig}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{m}}{\bid{i}}{\dirStop}{\emptyTrainProg}\}}{\regulator}{\switches}{\signals}}
	}
\end{mathpar}

The last two rules for trains describe what happens when a train actually moves, i.e. when its position changes. %Two cases can occur, depending on the list of intermediate blocks $\posFmt{N}$ in the current order of the train. 
 If $\posFmt{N}$ contains only the upcoming block, then upon leaving the block, the order is removed (rule \ruleRef{UntilNext}); otherwise the upcoming block is simply popped from $\posFmt{N}$ (rule \ruleRef{Until}). Notice that in both cases, the state of the guard changes to \guardS, meaning that signals will be updated immediately after that rule. Finally, when the train reaches intermediate positions (rule \ruleRef{Until}), a train event is enqueued for the regulator.

\begin{mathpar}
	\inferrule*[left=\ruleDef{Until}] {
		\suc{\bid{i}}{d}{\switches} = \bid{j} \neq \nosuc
		\and
		\signalF{\bid{i}}{d} = \siggreen
		\and
		\posFmt{N} = \bid{j} \cdot \posFmt{N'}
		\and
		\posFmt{N'} \neq \emptyList
	}{
		\redTuple{\guardT}{\bufTrain}{\bufSig}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{m}}{\bid{i}}{d}{\trainConcat{\su{d}{\posFmt{N}}}{P}}\}}{\regulator}{\switches}{\signals}} \\
		\reduces		
		\redTuple{\guardS}{\push{\bufTrain}{\tid{m}}}{\bufSig}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{m}}{\bid{j}}{d}{\trainConcat{\su{d}{\posFmt{N'}}}{P}}\}}{\regulator}{\switches}{\signals}}
	}
\end{mathpar}
